Empirical Bayes analysis for detection of gene heterosis in RNAseq data

### 2014-12-17 ebayes_rnaseq_heterosis.tex
through intro

Recently, \citep{ji2014estimation} introduced an approach to assess gene heterosis using micro-array data where the response is assumed to be continuous. They built a hierarchical model for gene expression that allows borrowing of information across genes for means and variances. They introduced an empirical Bayes framework that first estimates model hyperparameters, then estimates the posterior distribution for gene-specific parameters, and finally computes heterosis probabilities based on integrals of regions under this posterior.

%\section{Heterosis}
%\label{s:heterosis}

% Well before heterosis was scientifically described by , humans had been using heterosis for various practical purposes. Within the last century, heterosis has been used to improve many species for food, feed, and fuel industries. A few relatively recent examples include rice , alfalfa , tomatoes . Despite the intensive study and successful utilization of heterosis, the basic genomic mechanisms responsible for heterosis remain unclear .
% 
% \cite{shull1952beginnings} coined the term heterosis and defined it as 
% \begin{quote}
% the interpretation of increased vigor, size, fruitfulness, speed of development, resistance to disease and to insect pests, or to climatic rigors of any kind, manifested by crossbred organisms as compared with corresponding inbreds, as the specific results of unlikeness in the constitutions of the uniting parental gametes. 
% \end{quote}
% 
% For example, the maize hybrid produced from a cross of parental inbred lines B73 and Mo17 is larger in size, has lower maturation time, and has higher grain yield than both of its parents \cite{hallauer2010quantitative}. Data from \citep{paschold2012complementation}
% 
% \citep{ji2014estimation}

@@

An RNAseq experiment to identify genes demonstrating heterosis consists of three genetic varieties: two parental varieties (P1 and P2) and a cross between these two varieties called the hybrid (H). For each variety, replicate RNA samples are isolated and assessed for quality. cDNA libraries, i.e. short cDNA fragments, are constructed and next generation sequencing technology is used to determine the \emph{reads}, i.e. nucleotide sequences in the cDNA libraries. These reads are processed using bioinformatic algorithms to match the reads to genes, or specific gene transcripts, exons, microRNAs, etc. The result of matching reads to features are summarized by a gene $\times$ sample matrix of counts. For more details on RNAseq experiments see \cite{nettleton2014design} and for biological background on the details of these steps see \cite{paschold2012complementation}.

@@

To perform the optimization, we use the statistical software {\tt Stan} \citep{stan-software:2014} run through the RStan interface \citep{rstan-software:2014} in R \citep{R2014}. The Stan software allows construction of general Bayesian models with the primary restriction that all unknown parameters must be continuous. Once a model is constructed, the software can be used to obtain MAP estimates \cite[Section 50.3,][]{stan-manual:2014} using the {\tt optimizing} function in RStan. We found it necessary to use the L-BFGS algorithm \cite[Section 55,][]{stan-manual:2014} presumably due to the large dimension of the optimization. The {\tt Stan} software can also be used to run a Markov chain Monte Carlo (MCMC) algorithm to obtain samples from the posterior using the {\tt sampling} function in RStan. By default, Stan uses a variant of Hamiltonion Monte Carlo \citep{neal2011mcmc} called the no-U-turn sampler \citep{hoffman2013no}.

>>>>

Recently, \cite{ji2014estimation} introduced an approach to assess gene heterosis using microarray data where the data are assumed to be continuous. They built a normal hierarchical model for gene expression that allows borrowing of information across genes for means and variances. They introduced an empirical Bayes framework that first estimates model hyperparameters, then estimates the posterior distribution for gene-specific parameters conditional on those hyperparameters, and finally computes heterosis probabilities based on integrals of regions under this posterior. This development was necessary due to the composite null hypothesis when testing for heterosis making many available methods ill-suited to heterosis identification. When transitioning from continuous to count data, the difficulty in a composite null hypothesis remains. We following \citeauthor{ji2014estimation} in constructing a hierarchical model now built off a negative-binomial rather than normal data model. We also utilize an empirical Bayes approach based on obtaining \emph{maximum a posterior} estimates of hyperparameters and posterior distribution for gene-specific parameters conditional on those hyperparameters.

@@

An RNA sequencing (RNAseq) experiment to identify genes demonstrating heterosis consists of three genetic varieties: two parental varieties (P1 and P2) and a cross between these two varieties called the hybrid (H). For each variety, replicate RNA samples are isolated and assessed for quality. cDNA libraries, i.e. short cDNA fragments, are constructed and next generation sequencing technology is used to determine the \emph{reads}, i.e. nucleotide sequences in the cDNA libraries. These reads are processed using bioinformatic algorithms to match the reads to genes, or specific gene transcripts, exons, microRNAs, etc. The result of matching reads to features are summarized by a gene $\times$ sample matrix of counts. For more details on RNAseq experiments see \cite{nettleton2014design} and for biological background on the details of these steps see \cite{paschold2012complementation}.

@@

To perform the optimization, we use the statistical software {\tt Stan} \citep{stan-software:2014} run through the RStan interface \citep{rstan-software:2014} in R \citep{R2014}. The Stan software allows construction of general Bayesian models with the primary restriction that all unknown parameters must be continuous. Once a model is constructed, the software can be used to obtain MAP estimates \cite[Section 50.3,][]{stan-manual:2014} using the {\tt optimizing} function in RStan. We found it necessary to use the L-BFGS algorithm \cite[Section 55,][]{stan-manual:2014} and reasonable starting values obtained via {\tt edgeR} presumably due to the large dimension of the optimization. The {\tt Stan} software can also be used to run a Markov chain Monte Carlo (MCMC) algorithm to obtain samples from the posterior using the {\tt sampling} function in RStan. By default, Stan uses a variant of Hamiltonion Monte Carlo \citep{neal2011mcmc} called the no-U-turn sampler \citep{hoffman2013no}.
